---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```
`oceanwaves` provides a set of functions to calculate summary statistics for

ocean waves, using a record of surface heights as input. For sea surface heights
derived from bottom-mounted pressure transducers, the package also contains a
function `prCorr` to correct for depth attenuation of the pressure signal. 

`waveStatsSP` produces wave height and period statistics using spectral analysis
methods, while `waveStatsZC` calculates additional wave height and period
statistics based on a zero-crossing algorithm. 

Pressure corrections and wave statistics functions were adapted from Urs
Neumeier's `waves` functions for MATLAB.
[http://neumeier.perso.ch/matlab/waves.html](http://neumeier.perso.ch/matlab/waves.html)  

To install this package from within R, first install the package `devtools`
[https://CRAN.R-project.org/package=devtools]{https://CRAN.R-project.org/package=devtools}
and then install this package from Github:

```{r install_github, eval = FALSE}
install.packages('devtools')
library(devtools)
install_github('millerlp/oceanwaves')
```

