---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```
`oceanwaves` provides a set of functions to calculate summary
statistics
for ocean waves, using a record of sea surface elevation as input. For sea surface
elevations
derived from bottom-mounted pressure transducers, the package also
contains a
function `prCorr()` to correct for depth attenuation of the pressure signal,
and the `swDepth()` function from the package `oce` can be used to convert pressure data into ocean surface elevations (see the included package vignette). 

`waveStatsSP()` produces wave height and period statistics using spectral analysis
methods, while `waveStatsZC()` calculates additional wave height and period
statistics based on a zero-crossing algorithm. 

See the package vignette for example workflows to proceed from
raw pressure data to summary wave statistics.

Pressure corrections and wave statistics functions were adapted from Urs
Neumeier's `waves` functions for MATLAB, developed from earlier work 
by Travis Mason and Magali Lecouturier.
[http://neumeier.perso.ch/matlab/waves.html](http://neumeier.perso.ch/matlab/waves.html)  

To install the development version of this package from within R, first install the package `devtools`
[https://CRAN.R-project.org/package=devtools](https://CRAN.R-project.org/package=devtools)
and then install this package from Github:

```{r install_github, eval = FALSE}
install.packages('devtools')
library(devtools)
install_github('millerlp/oceanwaves')
```

